### 2016-12-12 paper/chapters/vae.tex
remove theta from prior distribution
sbn + VI + draw

We use VAEs when we have a complicated distribution $p_\theta(x,z)$ with unknown latent variables $z$.
The prior distribution $p_\theta(z)$ over the latent structure is a centered isotropic(?) multivariate Gaussian denoted by $\mathcal{N}(z;0, I)$.
The constructed architecture uses an probabilistic encoder $q_\theta(z|x)$ and a probabilistic decoder $p_\theta(x|z)$ in a form of neural networks, in the original paper MLP are used\cite{vae:2013} but there are various extensions (see \ref{sub:vae_extensions}, TODO).
Because the posterior is intractable ($p_\theta(x)$), we instead approximately maximize the lower variational bound $L(\theta,\phi;x)$.\\
\begin{equation}
  \mathcal{L}(\theta,\phi;x) = -D_{KL}(q_\theta(z|x)||p_\theta(z)) + \mathbb{E}_{q_\theta(z|x)}[log_{p_\theta}(x|z)]
\end{equation}

@@

\subsection{Inference}
\label{sub:vae_inference}
\begin{figure}[htb]
\centering
%\resizebox{5cm}{!}{\input{media/vae_architecture.tikz}}
\includegraphics{media/manifold.pdf}
  \caption{MNIST 2-dimensional manifold}\label{fig:vae_manifold}
  \medskip
  \small
  Manifold learned by a VAE using a 2-dimensional latent space for $z \in [-3,3]^2$.
\end{figure}

>>>>

We use VAEs when we have a complicated distribution $p_\theta(x,z)$ with unknown latent variables $z$.
The prior distribution $p(z)$ over the latent structure is a centered isotropic(?) multivariate Gaussian denoted by $\mathcal{N}(z;0, I)$.
The constructed architecture uses an probabilistic encoder $q_\theta(z|x)$ and a probabilistic decoder $p_\theta(x|z)$ in a form of neural networks, in the original paper MLP are used\cite{vae:2013} but there are various extensions (see \ref{sub:vae_extensions}, TODO).
Because the posterior is intractable ($p_\theta(x)$), we instead approximately maximize the lower variational bound $L(\theta,\phi;x)$.\\
\begin{equation}
  \mathcal{L}(\theta,\phi;x) = -D_{KL}(q_\theta(z|x)||p(z)) + \mathbb{E}_{q_\theta(z|x)}[log_{p_\theta}(x|z)]
\end{equation}

@@

\subsection{Inference}
\label{sub:vae_inference}
In bayesian analysis, statistical inference is the process of evaluating $p(z|x)$ for a given conditional probability $p(x|z)$.
% more info on marginal distribution, posterior inference etc!
There are two approaches to performing this task, Markov chain Monte Carlo (MCMC) methods using Gibbs sampling and variational inference.
MCMC methods have the advantage of being asympotically exact and unbiased while variational methods are computationally faster but biased and not guaranteed to approaching the distribution.
In the case of VAE, variational inference is also used in order to backpropagate through the layers (using the reparameterization trick).

\begin{figure}[htb]
\centering
%\resizebox{5cm}{!}{\input{media/vae_architecture.tikz}}
\includegraphics{media/manifold.pdf}
  \caption{MNIST 2-dimensional manifold}\label{fig:vae_manifold}
  \medskip
  \small
  Manifold learned by a VAE using a 2-dimensional latent space for $z \in [-3,3]^2$.
\end{figure}

### 2016-12-12 paper/chapters/vae/iwae.tex
sbn + VI + draw

Importance Weighted Autoencoders (IWAE) extends on the VAE framework using $k$-sample importance weighting of the log-likelihood.
The authors of the papers argue that using more samples will only tighten the lower variational bound.
$$
\log p(x) \geq \mathcal{L}_{k+1}(x) \geq \mathcal{L}_{k}(x)
$$

>>>>

% write about multiple stochastic layers
Until now, the presented model includes only one stochastic layer.
\cite{dlgm:2014} have presented a model for learning through multiple stochastic layers which can be easily applied to the VAE framework as shown in \ref{fig:vae_multiple_layers}.

\begin{figure}[htb]
\centering
\includestandalone[width=\linewidth,mode=buildnew]{media/vae_multiple_layers}
  \caption{VAE with multiple stochastic layers }\label{fig:vae_multiple_layers}
  \medskip
  \small
  The left upward arrows indicate the recognition networks $q(h_1|x)$ and $q(h_2|h_1)$.
  Note that the recognition networks can have multiple deterministic layers (including non-linearities).\\

  $p(h_2)$ denotes the probability distribution over the highest(?) latent space.\\

  On the right, the two conditional distributions $p(h_1|h_2)$ and $p(x|h_2)$ are representing two networks used for the generative process.
\end{figure}
% write about k passes (just averages)

% write about IWAE, extends these k passes with importance weights
Importance Weighted Autoencoders (IWAE) extends on the VAE framework using $k$-sample importance weighting of the log-likelihood.
% add intuition for IWAE, ruslan explained it very well during his DL summer school talk! (talk_Montreal_2016_Salakhutdinov.pdf)
Using importance weighting can be informally interpreted as putting more weight on data with high likelihood.

The authors of the papers argue that using more samples will only tighten the lower variational bound.
$$
\log p(x) \geq \mathcal{L}_{k+1}(x) \geq \mathcal{L}_{k}(x)
$$
